---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r load_data}
data <- read.csv(unzip('activity.zip'),colClasses = c("numeric", "character", "numeric"))
```

## What is mean total number of steps taken per day?
```{r mean_total_number_of_steps}
# Ignore missing values
data_NA_rm <- data[which(!is.na(data$steps)),]

# Calculate the total number of steps taken per day
steps_per_day <- aggregate(steps ~ date, data_NA_rm, sum)

# Histogram of the total number of steps taken each day
hist(steps_per_day$steps, main = "Total steps per day", xlab = "day")

#Calculate the mean and median of the total number of steps taken per day
steps_mean <- prettyNum(round(mean(steps_per_day$steps),2))
steps_median <- prettyNum(median(steps_per_day$steps))
```

The mean of the total number of steps taken per day is `r steps_mean`, the median is `r steps_median`.

## What is the average daily activity pattern?
```{r average_daily_activity_pattern}
# Calculate average number of steps taken in each 5-minute interval, averaged across all days
steps_per_interval <- aggregate(steps ~ interval, data_NA_rm, mean)

# Time series plot of the interval and the average number of steps taken, averaged across all days
plot(x=steps_per_interval$interval,y=steps_per_interval$steps,type='l', 
     main="Average steps per interval",xlab='interval', ylab='steps')
     
# Calculate interval, that contains the maximum number of steps (on average across all the days)
max_interval <- steps_per_interval$interval[steps_per_interval$steps == max(steps_per_interval$steps)]
```

The 5-minute interval, which on average across all the days contains the maximum number of steps is `r max_interval`.


## Imputing missing values

My strategy for filling in all of the missing values in the dataset is to add the mean for that 5-minute interval.

```{r missing_values}
# Calculate and report the total number of missing values in the dataset
no_NA <- sum(is.na(data$steps))

# Create a new dataset that is equal to the original dataset but with the missing data filled in by adding the mean for that 5-minute interval.
data_filled <- data
for (i in 1: nrow(data_filled)){
  if(is.na(data_filled[i,'steps'])){
    data_filled[i,'steps'] <- steps_per_interval$steps[steps_per_interval$interval == data_filled[i,'interval']]
  }
}

# Calculate average number of steps taken in each 5-minute interval, averaged across all days
steps_per_day_filled <- aggregate(steps ~ date, data_filled, sum)

#Make a histogram of the total number of steps taken each day. 
hist(steps_per_day_filled$steps, main = "Total steps per day (NAs filled)", xlab = "day")

#Calculate the mean and median of the total number of steps taken per day
steps_mean_filled <- prettyNum(round(mean(steps_per_day_filled$steps),2))
steps_median_filled <- prettyNum(round(median(steps_per_day_filled$steps),2))
diff_means <- as.numeric(steps_mean_filled) - as.numeric(steps_mean)
diff_medians <- as.numeric(steps_median_filled) - as.numeric(steps_median)
```

The dataset contains `r no_NA` missing values. After filling the NAs, the mean of the total number of steps taken per day is `r steps_mean_filled`, the median is `r steps_median_filled`. The difference between the filled and unfilled means is `r diff_means`, the difference between the medians is `r diff_medians`.

## Are there differences in activity patterns between weekdays and weekends?
```{r differences_weekdays_weekends, fig.height=8}
#Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
data_filled$day <- weekdays(as.Date(data$date))
data_filled$day_type <- 'weekday'
data_filled$day_type[data_filled$day=='Samstag'] <- 'weekend'
data_filled$day_type[data_filled$day=='Sonntag'] <- 'weekend'
data_filled$day_type <- as.factor(data_filled$day_type)

#Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
steps_per_interval_filled <- data.frame(steps_weekend=NA,steps_weekday=NA,interval=unique(data$interval))
steps_per_interval_filled$interval <- as.character(steps_per_interval_filled$interval)
for (interval_i in 1:nrow(steps_per_interval_filled)){   
  steps_per_interval_filled[interval_i,'steps_weekend'] <- 
    mean(data_filled[which(data_filled$interval == steps_per_interval_filled[interval_i,'interval'] &
                             data_filled$day_type == 'weekend'),'steps'])
  steps_per_interval_filled[interval_i,'steps_weekday'] <- 
    mean(data_filled[which(data_filled$interval == steps_per_interval_filled[interval_i,'interval'] &
                             data_filled$day_type == 'weekday'),'steps'])
}
# The assignment said, "feel free to use any plotting system in R", so I keep using the base system and create the panel equivalent
par(mfrow=c(2,1))
plot(x=steps_per_interval_filled$interval,y=steps_per_interval_filled$steps_weekday,type='l', 
     main="Average steps per interval (weekdays)",xlab='interval', ylab='steps')
plot(x=steps_per_interval_filled$interval,y=steps_per_interval_filled$steps_weekend,type='l', 
     main="Average steps per interval (weekend)",xlab='interval', ylab='steps')
```
